Effects of bariatric surgery on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis

### 2021-03-22 manuscript/draft_rsfmri.Rmd

---
title: "Effects of bariatric surgery on functional connectivity of the reward and default mode network: a pre-registered analysis"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Evelyn Medawar [1]
- Kristin Prehn [4][5]
- Jürgen Ordemann [8][9]
- Agnes Floel [4][5]
- A. Veronica Witte [3][1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Clinic of Cognitive Neurology, University Clinic Leipzig \newline 4 Department of Neurology and NeuroCure Clinical Research Center, Charité University Medicine, Berlin \newline 5 Department of Psychology, Medical School Hamburg, Hamburg \newline 6 Department of Neurology, University of Greifswald, Greifswald \newline 7 German Center for Neurodegenerative Diseases, Standort Rostock/Greifswald, Greifswald \newline 8 Center for Bariatric and Metabolic Surgery, Charité University Medicine, Berlin \newline 9 Center for Bariatric and Metabolic Surgery, Vivantes Clinic Spandau, Berlin
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{setspace}
- \onehalfspacing
indent: true
---

@@

Obesity imposes serious health risks and involves alterations in resting-state functional connectivity  of brain networks involved in eating behavior. Bariatric surgery is an effective treatment, but its effects on functional connectivity are still under debate. In this pre-registered study, we investigated the effects of bariatric surgery on major resting-state brain networks (reward and default mode network) in a longitudinal controlled design. 
33 bariatric surgery patients and 15 obese waiting-list control patients (37 females; aged 44.15 ± 11.86 SD years (range 21-68)) underwent magnetic resonance imaging at baseline, after 6 and 12 months. We conducted a pre-registered whole-brain time-by-group interaction analysis, and a time-by-group interaction analysis on within-network connectivity (https://osf.io/f8tpn/, https://osf.io/59bh7/). In exploratory analyses, we investigated the effects of weight loss and head motion. 
Bariatric surgery compared to waiting did not significantly affect functional connectivity (FWE-corrected $p > .05$), neither whole-brain nor within-network. In exploratory analyses, surgery-related BMI decrease (FWE-corrected $p = .041$) and higher average head motion (FWE-corrected $p = .021$) resulted in significantly stronger connectivity of the reward network with medial posterior frontal regions. 
This pre-registered well-controlled study did not support a strong effect of bariatric surgery, compared to waiting, on major resting-state brain networks after 6 months. Exploratory analyses indicated that head motion might have confounded the effects. Data pooling and more rigorous control of within-scanner head motion during data acquisition are needed to substantiate effects of bariatric surgery on brain organization.

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Excess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life [@Albanese_2017]. Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI ($> 35$ kg/m$^2$). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical procedure, but other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) are also common [@Berthoud_2011].\
Bariatric surgery leads to weight loss due to malabsorption of nutrients, reduced digestion efficiency and altered food perception, appetite and central regulation of food intake [@Mulla_2017; @Brutman_2019]. Precise mechanisms how bariatric surgery leads to altered appetitive signaling are yet to be elucidated. One option to adress these questions on brain-behavior relationships is to use resting-state functional magnetic resonance imaging (rsfMRI), a technique capturing the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.\
The RN comprises the ventromedial prefrontal cortex (vmPFC), the nucleus accumbens (NAcc), the putamen, the amygdala and the anterior insula  [@Odoherty_2004;@Liu_2011]. These brain regions have been suggested to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. RsfMRI studies also showed increased local FC of reward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018].\
The DMN includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. After bariatric surgery, a normalization of the connectivity between DMN and cognitive control and salience brain regions might occur, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].\
In sum, while there is some evidence hinting to a role for DMN and RN FC in altered regulation of food intake after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohorts of patients, without adequate obese control groups, and did not rigorously separate confirmatory from exploratory analyses [@George_2016].\
Further, while higher BMI has been consistently associated with more head motion during rsfMRI [@Beyer_2017;@Hodgson_2016], previous studies in bariatric surgery patients have not taken this important confounder of FC into account. In the present sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Thus, we aimed to rigorously control for motion-related variance in our analyses.\

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité University Medicine Berlin, Germany. For more details see [@Prehn_2020]. We used all data acquired until April 2019.\
The study protocol was in accordance with the Helsinki Declaration and approved by the ethics committee of the Charité University Medicine Berlin. The study was registered at clinicaltrials.gov as NCT01554228. We pre-registered the present resting-state analyses on the OSF (https://osf.io/yp42s). We made additional changes (see https://osf.io/59bh7/) to the pre-registration after preprocessing the rsfMRI data as we realized some aspects of the analysis were inadequately described in the initial pre-registration. For a comparison of the pre-registration and the manuscript, please visit https://osf.io/45n9f/.\
Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (defined as MMSE score < $34$).\
In total, $51$ participants out of the originally enrolled $69$ subjects received MRI. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below), which led to a final data set with $101$ rsfMRI sessions.\
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`)). `r round(sum(subj_df$arterial_hypotension == 1, na.rm = T) / nrow(subj_df), 3)*100`% of participants had clinically diagnosed hypertension, `r round((sum(subj_df$diabetes_type_II == 1, na.rm = T) + sum(subj_df$diabetes_type_I == 1, na.rm = T))  / nrow(subj_df), 3)*100`% had type-2 or type-1 diabetes, and `r round(sum(subj_df$smoking_status == 1, na.rm = T) / nrow(subj_df), 3)*100`% reported to smoke.\
Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` females) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` females), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. 
Analyses were performed on all participants who provided at least one data point of rsfMRI data. `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point (for more details see Figure 1 in supplementary information (SI)). The pre-registered analysis of changes from baseline to followup included $24$ participants with both time points, in total $72$ data points.\
`r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="RYGB" & !is.na(subj_df$interv),])` patients underwent RYGB, `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="VSG" & !is.na(subj_df$interv),])` underwent VSG and `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="GB" & !is.na(subj_df$interv),])` GB, for `r nrow(subj_df[subj_df$condition=="IG" & is.na(subj_df$interv),])` patients in the intervention group information was not available. \
Participants arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after performing a psychological test battery (for details see @Prehn_2020).

@@

### Minimal preprocessing
Imaging data analysis was conducted using AFNI 19.1.05, ANTS 2.3.1, FSL 6.0.1 and FreeSurfer 6.0.0p1, wrapped in a nipype workflow (version 1.2.0) in Python 2.7.15 which can be found on https://github.com/fBeyer89/ADI_preproc/.
T1-weighted images were first processed by FreeSurfer's cross-sectional pipeline [@fs_cross]. Then, Freesurfer's longitudinal stream was applied to all cross-sectional runs [@fs_long]. Here, white matter (WM) and cerebral spinal fluid (CSF) masks were derived based on FreeSurfer's `aseg.mgz` segmentation file for quality control of rsfMRI preprocessing.  
The skull-stripped brain (`brainmask.mgz`) was then coregistered to the MNI152 1x1x1x mm template using ANTS [@avants2009ants]. 
Minimal functional preprocessing included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step. 
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).

### Denoising pipelines
In the pre-registration, we specified two denoising pipelines (AROMA+CC including ICA-AROMA + CompCor and AROMA+CC+GSR including ICA-AROMA + CompCor + global signal regression (GSR)) [@Parkes_2018; @Ciric_2017, for details see SI].

@@

**Anatomical quality control**
FreeSurfer cross-sectional and longitudinal  segmentations were visually checked according to @Klapwijk_2019. We excluded 5 datasets from 3 participants because of excessive head motion leading to failed pial reconstruction and anatomical-functional coregistration.\
<!--ADI009_fu2, ADI063_bl, ADI063_fuADI116_bl, ADI116_fu, ADI116_fu2-->
**RsfMRI quality control**
RsfMRI quality control was performed according to the protocol by @Ciric_2018 (see SI for more details). 
Head motion was quantified using mean framewise displacement (mFD) according to @Power_2012, and log-transformed for further analysis (logmFD).
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020].

## Functional connectivity (FC)
To derive RN and DMN FC maps, we used NAcc and PCC/precuneus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Based on FreeSurfer's aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLabelsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each preprocessing pipeline (minimally preprocessed,AROMA, AROMA+CC, AROMA+CC+GSR). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.

## Aggregated within-network FC 
To extract aggregated within-network FC, we first calculated the mean DMN and RN over all participants and time points, adjusted for age and sex. We used GSR-denoised data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average GSR-denoised FC from these masks.

@@

Statistical analyses were performed in MATLAB version 9.7.0.1190202 (R2019b, @MATLAB2018) using the SwE toolbox version 2.2.2 [@Guillaume_2014] as implemented in the Statistical Parametric Mapping software (SPM12.7770, Ashburner et al. -@spm12). The marginal model implemented in the SwE toolbox implicitly accounts for random effects without the need to specify them through the error term. We used a modified SwE assuming different covariance structures for the intervention and the control group because of their unbalanced sample size. We used an explicit brain mask, derived from the MNI ICBM “152 nonlinear 6th generation” atlas (re-sampled to 3 x 3 x 3 mm$^3$ and thresholded at 0.5 GM probability) for all analyses. Statistical analyses on the aggregated FC and imputation of missing data were performed in R version 3.6.1 [@team2013r].

<!-- MODEL SETUP --> 
### Confirmatory analyses
**Whole-brain analysis**
We tested the pre-registered hypothesis of a time-by-group interaction:  `FC=group + time + group * time`.\
As pre-registered, we performed the analysis for baseline and followup time points only, for AROMA+CC and AROMA+CC+GSR denoising pipelines and adjusting for no confounders (model CA1), age, sex and average logmFD of both timepoints (model CA2) and age, sex, average logmFD and baseline BMI (model CA3). 
Because the information about the BMI at baseline of one participant in the intervention group was missing, we employed multivariate imputation using the package `mice` based on subject identification, condition, baseline age, and available BMI values. This package allows taking into account the multi-level structure of the data and averaging across multiple possible values, in this case 50 different values, each derived from 10 iterations.\
**Aggregated FC analysis**
We analyzed the aggregate FC using linear mixed models in the R package lme4 [@Bates_2007]. We deviated from the pre-registration by only investigating data from the AROMA+CC+GSR denoising pipeline. First, we investigated the time-by-group interaction for baseline and followup time points only. We either performed no correction (model CA1), adjusted for baseline age, sex, average of logmFD of both time points (model CA2) or additionally adjusted for baseline BMI (model CA3). We performed model comparison between R1 and R0 models, where $R1 = lmer(aggFC ~ timepoint*group + age + sex + (1|subj))$, and $R0 = lmer(aggFC ~ timepoint + group +  age + sex + (1|subj))$. As specified in the pre-registration, we repeated the above-mentioned interaction analysis for all three time points.

>>>>

---
title: "Effects of bariatric surgery on resting-state functional connectivity of the reward and default mode network: a pre-registered analysis"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Evelyn Medawar [1]
- Kristin Prehn [4][5]
- Jürgen Ordemann [8][9]
- Agnes Floel [4][5]
- A. Veronica Witte [3][1][2]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3 Clinic of Cognitive Neurology, University Clinic Leipzig \newline 4 Department of Neurology and NeuroCure Clinical Research Center, Charité University Medicine, Berlin \newline 5 Department of Psychology, Medical School Hamburg, Hamburg \newline 6 Department of Neurology, University of Greifswald, Greifswald \newline 7 German Center for Neurodegenerative Diseases, Standort Rostock/Greifswald, Greifswald \newline 8 Center for Bariatric and Metabolic Surgery, Charité University Medicine, Berlin \newline 9 Center for Bariatric and Metabolic Surgery, Vivantes Clinic Spandau, Berlin
output:
  bookdown::pdf_book:
    extra_dependencies: "subfig"
    fig_caption: yes
    number_sections: yes
    toc: yes
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{setspace}
- \onehalfspacing
indent: true
---

@@

Obesity imposes serious health risks and relates to differences in resting-state functional connectivity (FC) of brain networks involved in eating behavior. Bariatric surgery is an effective treatment, but its effects on FC are still under debate. In this pre-registered analysis, we aimed to determine the effects of bariatric surgery on FC in the reward (RN) and default mode network (DMN). 
33 bariatric surgery patients and 15 obese waiting-list control patients (37 females; aged 44.15 ± 11.86 SD years (range 21-68)) underwent magnetic resonance imaging at baseline, after 6 and 12 months. We pre-registered (https://osf.io/f8tpn/https://osf.io/59bh7/) a whole-brain time-by-group interaction analysis, and a time-by-group interaction analysis on within-network FC. In exploratory analyses, we investigated all three time points and assessed the effects of weight loss and head motion. 
Bariatric surgery compared to waiting did not significantly affect FC of the RN and DMN (FWE p > .05). There also was no interaction effect on within-network FC. In exploratory analyses, surgery-related BMI decrease (FWE p= .041) and higher average head motion (FWE p= .021) resulted in significantly stronger FC in medial posterior frontal regions within the RN. 
This pre-registered analysis of longitudinal neuroimaging data did not support a strong effect of bariatric surgery, compared to waiting, on RN and DMN FC after 6 months. Exploratory analyses indicated that differences in head motion might have confounded the effects, thus a more rigorous control of within-scanner head motion during data acquisition is needed to better understand the effects of bariatric surgery on brain organization.

@@

Obesity is a worldwide health issue, entailing huge personal and societal costs. Excess amount of body fat not only affects cardiovascular and metabolic health, but also increases the risk for cognitive decline and dementia later in life [@Albanese_2017]. Conservative treatment options including behavioral therapy often do not yield the desired weight loss, especially in patients with very high BMI ($> 35$ kg/m$^2$). Here, bariatric surgery is a viable option to rapidly induce weight loss and improve glycemic status. RYBG is the most invasive, yet most effective surgical procedure, but other techniques like vertical sleeve gastrectomy (VSG) and gastric banding (GB) are also common [@Berthoud_2011]. 
Bariatric surgery leads to weight loss due to malabsorption of nutrients, reduced digestion efficiency and altered food perception, appetite and central regulation of food intake [@Mulla_2017; @Brutman_2019]. Precise mechanisms how bariatric surgery leads to altered appetitive signaling are yet to be elucidated. One option to adress these questions on brain-behavior relationships is to use resting-state functional magnetic resonance imaging (rsfMRI), a technique capturing the dynamic organization of the brain. Functional connectivity (FC) networks, i.e. brain regions with correlated neural activity over time, are in anatomical correspondence with specific brain networks involved in cognitive processes, including attention and executive control [@Smith_2009]. The reward network (RN), processing hedonic value and internal motivation, and the default mode network (DMN), a higher-order network, involved in interoception and governing shifts between external-internal processes,  are promising candidates to mediate altered central regulation of food intake after bariatric surgery.
The RN (RN) comprises the ventromedial prefrontal cortex (vmPFC), the nucleus accumbens (NAcc), the putamen, the amygdala and the anterior insula  [@Odoherty_2004;@Liu_2011]. These brain regions have been suggested to guide food valuation processes and decision-making in humans [@Bartra_2013; @Hare_2011; @Hutcherson_2012; @Schmidt_2018]. Frequently, obesity has been associated with hyperactivation of RN regions during anticipation of (high-caloric) food cues, and in contrast, reduced activation to actual taste of these foods [@Devoto_2018; @Garcia_2014; @Meng_2020; @Stoeckel_2009; though see @Morys_2020]. RsfMRI studies also showed increased local FC of reward network regions, i.e. NAcc, vmPFC, putamen, insula [@Contreras_2017; @Coveleskie_2015;@Hogenkamp_2016], and altered connectivity with salience, homeostatic and sensorimotor networks [@Lips_2014;  @Wijngaarden_2015]. In bariatric surgery patients, connectivity  within the RN (e.g. putamen, OFC) might be normalized by the surgery, however the evidence is limited due to a lack of longitudinal obese control groups  [@Duan_2020; @Wiemerslage_2016]. Possibly, a reconfiguration of the fronto-striatal brain networks could emerge from altered gut signaling , e.g. changes in ghrelin levels, via hypothalamic-striatal projections [@Li_2019; @Karra_2013; but @Zoon_2018]. 
The DMN includes the posterior cingulate cortex (PCC)/precuneus, the medial prefrontal cortex (mPFC) and the inferior and lateral parietal cortex [@Raichle_2015]. Higher BMI and obesity have been associated with a pattern of decreased within DMN FC and increased FC of DMN regions to other networks, i.e. salience and sensory networks in several resting-state and task-based rsfMRI studies [@Beyer_2017; @Kullmann_2011; @Chao_2018; @Wijngaarden_2015; @Borowitz_2020; @Ding_2020; @Doucet_2017; @Sadler_2018]. After bariatric surgery, a normalization of the connectivity between DMN and cognitive control and salience brain regions might occur, yet no study has included a longitudinal control group [@Li_2018; @Olivo_2017; @Frank_2013].
In sum, while there is some evidence hinting to a role for DMN and RN FC in altered regulation of food intake after bariatric surgery, the existing evidence is inconclusive. Most studies have investigated small cohorts of patients, without adequate obese control groups, and did not rigorously separate confirmatory from exploratory analyses [@George_2016].
Further, while higher BMI has been consistently associated with more head motion during rsfMRI [@Beyer_2017;@Hodgson_2016], previous studies in bariatric surgery patients have not taken this important confounder of FC into account. In the present sample, we previously reported a group-by-time interaction on head motion [@Beyer_2020]. Thus, we aimed to rigorously control for motion-related variance in our analyses. 

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study investigated the effects of bariatric surgery on brain structure and function in a prospective design at the Charité University Medicine Berlin, Germany. For more details see [@Prehn_2020]. We used all data acquired until April 2019.
The study protocol was in accordance with the Helsinki Declaration and approved by the ethics committee of the Charité University Medicine Berlin. The study was registered at clinicaltrials.gov as NCT01554228. We pre-registered the present resting-state analyses on the OSF https://osf.io/yp42s. We made additional changes https://osf.io/59bh7/ to the pre-registration after denoising as we realized some aspects of the analysis were inadequately described in the initial pre-registration.
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA -->

Participants were recruited from the Center for Bariatric and Metabolic Surgery at the Charité University Medicine Berlin. Inclusion criteria were, in accordance with German guidelines for bariatric surgery, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013]. Participants were aged between $18$ and $70$ years and had no history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (defined as MMSE score < $34$).
 In total, $51$ participants out of the originally enrolled $69$ subjects received MRI. Five data points of three subjects had to be excluded due to bad anatomical image quality (see below), which led to a final data set with $101$ rsfMRI sessions.  
<!-- DESCRIPTIVE STATISTICS -->
The final sample entailed `r nrow(subj_df)` morbidly obese individuals (`r nrow(filter(subj_df,Sex == "f"))` females; aged `r round(mean(subj_df$Age_BL),2)` $\pm$ `r round(sd(subj_df$Age_BL),2)` SD years (range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`)). `r round(sum(subj_df$arterial_hypotension == 1, na.rm = T) / nrow(subj_df), 3)*100`% of participants had clinically diagnosed hypertension, `r round((sum(subj_df$diabetes_type_II == 1, na.rm = T) + sum(subj_df$diabetes_type_I == 1, na.rm = T))  / nrow(subj_df), 3)*100`% had type-2 or type-1 diabetes, and `r round(sum(subj_df$smoking_status == 1, na.rm = T) / nrow(subj_df), 3)*100`% reported to smoke.
Participants either underwent surgery (n = `r cs["Sum","IG"]`, `r cs["f","IG"]` female) or were waiting list controls (n = `r cs["Sum","KG"]`, `r cs["f","KG"]` female), who waited for their health insurance's approval to undergo surgery. Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery/baseline appointment to capture both phases of rapid weight-loss and maintenance [@Maciejewski_2016]. 
Analyses were performed on all participants who provided at least one data point of rsfMRI data. `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for one time point (for more details see Table S1). The pre-registered analysis of changes from baseline to followup included $24$ participants with both time points, in total $72$ data points.

`r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="RYGB" & !is.na(subj_df$interv),])` patients underwent RYGB, `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="VSG" & !is.na(subj_df$interv),])` underwent VSG and `r nrow(subj_df[subj_df$condition=="IG" & subj_df$interv=="GB" & !is.na(subj_df$interv),])` GB, for `r nrow(subj_df[subj_df$condition=="IG" & is.na(subj_df$interv),])` patients in the intervention group information was not available.

Participants arrived in the morning (between 07:00 and 12:00 a.m.) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one hour break for breakfast. MRI scanning was done after performing a psychological test battery (for details see @Prehn_2020)).

@@

### Minimal preprocessing
Imaging data was conducted using AFNI 19.1.05, ANTS '2.3.1', FSL '6.0.1' and FreeSurfer '6.0.0p1', wrapped in a nipype workflow (version 1.2.0) in Python 2.7.15.
The preprocessing workflow included modules to process anatomical, functional and diffusion-weighted imaging in parallel. The workflow can be found on https://github.com/fBeyer89/ADI_preproc/.
T1-weighted images were first processed by FreeSurfer's cross-sectional pipeline [@fs_cross]. Then, Freesurfer's longitudinal stream was applied to all cross-sectional runs [@fs_long]. Here, white matter (WM) and cerebral spinal fluid (CSF) masks were derived based on FreeSurfer's `aseg.mgz` segmentation file for quality control of rsfMRI preprocessing.  
The skull-stripped brain (`brainmask.mgz`) was then coregistered to the MNI152 1x1x1x mm template using ANTS [@avants2009ants]. The following parameters were used for rigid [metric: mutual information, number of iterations: [1000, 500, 250, 100]], affine [metric: mutual information, number of iterations: [1000, 500, 250, 100]] and SyN: [metric: cross-correlation, iterations: [100, 70, 50, 20]] registration steps with shrink factors [8, 4, 2, 1], and smoothing sigmas=[3, 2, 1, 0] for each step.
Minimal functional preprocessing included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step. 
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).

### Denoising pipelines
We specified two denoising pipelines (ICA-AROMA + CompCor and ICA-AROMA + CompCor + global signal regression (GSR) in the pre-registration. AROMA+CC and AROMA+CC+GSR [@Parkes_2018; @Ciric_2017, for details see SI].

@@

**Anatomical quality control**
FreeSurfer cross-sectional and longitudinal  segmentations were visually checked according to @Klapwijk_2019. We excluded 5 datasets from 3 participants because of excessive head motion leading to failed pial reconstruction and anatomical-functional coregistration.
<!--ADI009_fu2, ADI063_bl, ADI063_fuADI116_bl, ADI116_fu, ADI116_fu2-->

**RsfMRI quality control**
RsfMRI quality control was performed according to the protocol by @Ciric_2018 (see SI for more details). 
We calculated mean framewise displacement (mFD) according to @Power_2012, and log-transformed it for further analysis (logmFD).
As pre-registered, we did not exclude anybody based on high average head motion [@Beyer_2020].

## Functional connectivity (FC)
To derive RN and DMN FC maps, we used NAcc and PCC/precuneus as seed regions of interest (ROI), respectively. We did not select vmPFC for the RN due to low SNR. Based on FreeSurfer's aseg.mgz and Desikan-Killiany parcellation, we created seed masks using mri_binarize (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used NiftiLabelsMasker and NiftiMasker to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson’s correlation between them with numpy.dot, performed r-to-z Fisher-transformation and saved the resulting correlation maps for each preprocessing pipeline (minimally preprocessed,AROMA, AROMA+CC, AROMA+GSR). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.

## Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and RN over all participants and NAccs, adjusted for age and sex. We used GSR regressed data as input and clusterwise bootstrapping with $N = 1000$. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of FWE-corrected $p < 0.05$. We extracted the average FC from these masks for GSR-data.

@@

Statistical analyses were performed in MATLAB version 9.7.0.1190202 (R2019b, @MATLAB2018) using the SwE toolbox version 2.2.2 [@Guillaume_2014] as implemented in the Statistical Parametric Mapping software (SPM12.7770, Ashburner et al. -@spm12). The marginal model implemented in the SwE toolbox implicitly accounts for random effects without the need to specify them through the error term. It therefore accommodates any repeated measurement covariance structures. We used a modified SwE assuming different covariance structures for the intervention and the control group because of their unbalanced sample size, so heteroscedasticity would be considered. We used an explicit brain mask, derived from the MNI ICBM “152 nonlinear 6th generation” atlas (re-sampled to 3 x 3 x 3 mm$^3$ and thresholded at 0.5 GM probability) for all analyses. Statistical analyses on the aggregated FC and imputation were performed in R version 3.6.1 [@team2013r].

<!-- MODEL SETUP --> 
### Confirmatory analyses
#### Whole-brain analysis
We tested the pre-registered hypothesis of a time-by-group interaction:  `FC=group + time + group * time`. 
As pre-registered, we performed the analysis for baseline and followup time points only, for AROMA+CC and AROMA+CC+GSR denoising pipelines and adjusting for no confounders (CA1), age, sex and average logmFD of both timepoints (CA2) and age, sex, average logmFD and baseline BMI (CA3). 
Because the information about the BMI at baseline of one participant in the intervention group was missing, we employed multivariate imputation using the package mice based on subject identification, condition, baseline age, and available BMI values. This package allows taking into account the multi-level structure of the data and averaging across multiple possible values, in this case 50 different values, each derived from 10 iterations.

#### Aggregated FC analysis
We analyzed the aggregate FC using linear mixed models in the R package lme4 [@Bates_2007]. We deviated from the pre-registration by only investigating data from the AROMA+CC+GSR denoising pipeline. First, we investigated the time-by-group interaction for baseline and followup time points only. We either performed no correction (CA1), adjusted for baseline age, sex, average of logmFD of both time points (CA2) or additionally adjusted for baseline BMI (CA3). We performed model comparison between R1 and R0 models, where $R1 = lmer(aggFC ~ timepoint*group + age + sex + (1|subj))$, and $R0 = lmer(aggFC ~ timepoint + group +  age + sex + (1|subj))$. As specified in the pre-registration, we repeated the above-mentioned interaction analysis for all three time points.

### Exploratory analysis (pre-registered)
#### Whole-brain analysis
As described in the pre-registration, we performed an exploratory analysis of the relationship of average and change in BMI in a separate group of models. As proposed by @Guillaume_2014, we calculated the between- and within subject centered values of BMI. This model, containing average BMI and BMI change, allowed us to disentangle the differential effects of these variables on FC. We first estimated their effects in a model adjusting for baseline age and sex (Model EA 1) and then additionally controlling for logmFD (Model EA 2). As we previously reported correlated change in BMI and mFD in this sample [@Beyer_2019], we explored a refined model including average BMI and logmFD and change in both measures, along with baseline age and sex (Model EA 3). Here, we aimed to see whether any effect of change in BMI would be detectable when adjusting for the change in head motion.

#### Aggregated FC analysis
We also performed an exploratory analysis with average BMI and change in BMI as predictors of the aggregated FC from AROMA+CC+GSR denoised data of both networks. We calculated three models with average and change in BMI as predictors of interest and adjusted for baseline age and sex (Model EA 1), logmFD (Model EA 2) and average and change in logmFD (Model EA 3).
